---
output: github_document
title: 'Wilcoxon-Mann-Whitney Test of No Group Discrimination (Continuous Variables)'
bibliography: inst/refs.bib
link-citations: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  warning = FALSE,
  message = FALSE
)
library(wmwAUC)
 if (!requireNamespace("gemR", quietly = TRUE)) {
     install.packages("gemR")
 }
library('gemR')
#
if (!requireNamespace("gss", quietly = TRUE)) {
  install.packages("gss")
}
library('gss')
#
if (!requireNamespace("VGAM", quietly = TRUE)) {
  install.packages("VGAM")
}
library('gss')
#
# 
```


# wmwAUC

<!-- badges: start -->
<!-- badges: end -->

Statistical theory incorrectly states that the Wilcoxon-Mann-Whitney (WMW) test examines
$\mathrm{H_0\colon F = G}$. We demonstrate this theoretical claim is untenable: 
the WMW statistic, when standardized, is the empirical AUC (eAUC), which measures $P(X < Y)$ 
and cannot detect distributional equality. Through Monte Carlo analysis of zero-mean 
heteroscedastic Gaussians and corresponding asymptotic theory, we show that WMW
tests $\mathrm{H_0\colon AUC = 0.5}$ (no systematic discrimination) rather than 
distributional equality. Moreover, the traditional alternative hypothesis of 
stochastic dominance is unnecessarily restrictive; WMW is consistent against the
broader alternative $\mathrm{H_1\colon AUC \neq 0.5}$, as established by Van Dantzig (1951).
We provide theoretical framework and implementation consistent with what the WMW eAUC 
statistic actually computes, including tie-corrected asymptotics and finite-sample 
bias corrections. For detalis, see [@grendar2025wmw].

The primary goal of wmwAUC is to provide inferences for the Wilcoxon-Mann-Whitney test
of $\mathrm{H_0\colon AUC = 0.5}$. 

## Installation

You can install the development version of wmwAUC using

``` r
devtools::install_github('grendar/wmwAUC')
```

## Simulation 1

Consider the setting of two zero-mean different-scale gaussians. Then the traditional
$\mathrm{H_0\colon F = G}$ of WMW test is false and $\mathrm{H_0\colon F \neq G}$ holds. 

The Monte Carlo simulation demonstrates that the normalized test statistic $U/(n_1n_2)$ which is just eAUC, concentrates asymptotically on 0.5 - the value expected under a true null hypothesis.

If WMW tested distributional equality, the test statistic should not concentrate on its null value when distributions clearly differ.

Also note that under $\mathrm{H_0\colon F \neq G}$, p-values should concentrate near zero, yet the observed distribution  is nearly uniform with a slightly elevated first bins, consistent with testing a true null hypothesis ($\mathrm{H_0\colon AUC = 0.5}$) using miscalibrated variance estimation.

```{r simulation1}
#############################################################################
#
# Simulation 1: H0: F=G is erroneously too broad
#
#############################################################################
#
# This simulation takes several minutes to complete
# N = 10000
# n = 1000
# set.seed(123L)
# pval_wt = pval_wmw = eauc = numeric(N)
# for (i in 1:N) {
#
#  x = rnorm(n, sd = 0.1)
#   y = rnorm(n, sd = 3)
#   # wilcox.test() of H0: F = G
#   wt = wilcox.test(x, y)
#   pval_wt[i] = wt$p.value
#   # wmw_test() of H0: AUC = 0.5
#   pval_wmw[i] = wmw_pvalue(x, y)
#   # eAUC
#   eauc[i] = wt$statistic/(n*n)
#   #
# }
data(simulation1)  # List eauc, pval_wt, pval_wmw
#
```

```{r, hist_sim1_1, fig.width=5, fig.height=5, include=T, fig.align='center', fig.show='hold', out.width = "50%", echo = FALSE}
hist(simulation1$eauc)
```

Empirical AUC centered at 0.5 despite $\mathrm{F \neq G}$.

```{r, hist_sim1_2, fig.width=5, fig.height=5, include=T, fig.align='center', fig.show='hold', out.width = "50%", echo = FALSE}
hist(simulation1$pval_wt) 
```

Traditional p-values under $\mathrm{H_1}$ should concentrate near 0.

```{r, hist_sim1_3, fig.width=5, fig.height=5, include=T, fig.align='center', fig.show='hold', out.width = "50%", echo = FALSE}
# Correct p-values (correctly testing AUC = 0.5)  
hist(simulation1$pval_wmw)
```

Correct p-values for testing $\mathrm{H_0\colon AUC = 0.5}$.


## Simulation 2

The two zero-mean different-scale gaussians setting does not satisfy the traditional $\mathrm{H_1}$ of the stochastic dominance. But, as proved by Van Dantzig in 1951, WMW is consistent for broader $\mathrm{H_1\colon AUC \neq 0.5}$.


```{r simulation2}
#############################################################################
#
# Simulation 2: H1: F stoch. dominates G is too narrow
#               WMW is consistent for broader H1: AUC != 0.5   
#
#############################################################################
#
#
# This simulation takes several minutes to complete
# N = 10000
# n = 1000
# set.seed(123L)
# pval_wt = pval_wmw = eauc = numeric(N)
# for (i in 1:N) {
# #
#   # gaussians with different location and scale
#   # does not satisfy stochastic dominance
#   x = rnorm(n, 0, sd = 0.1)
#   y = rnorm(n, 0.5, sd = 3)
#   # wilcox.test H0: F = G vs H1: (F stochastically dominates G) OR (G stochastically dominates F)
#   wt = wilcox.test(x, y)
#   pval_wt[i] = wt$p.value
#   # wmw_test H0: AUC = 0.5 vs H1: AUC neq 0.5
#   pval_wmw[i] = wmw_pvalue(x, y)
#   # eAUC
#   eauc[i] = wt$statistic/(n*n)
# #
# }
data(simulation2)  # List of eauc, pval_wt, pval_wmw
# WMW detects broader alternatives than traditional stochastic dominance
```

```{r,hist_sim2, fig.width=5, fig.height=5, include=T, fig.align='center', fig.show='hold', out.width = "50%", echo = FALSE}
hist(simulation2$eauc)
hist(simulation2$pval_wmw)
hist(simulation2$pval_wt)

```

## Simulation 3 

Confidence interval for the pseudomedian is obtained by inverting the test; 
see `pseudomedian_ci()` for implementation that handles the edge cases in the same 
way as `wilcox.test()`.

In this simulation study, N = 500 MC replicates are created, of 300 samples from 
the standard normal distribution and 300 samples from the Laplace distribution with
location = 0, scale = 1. Properties of 95% confidence intervals obtained under H0: AUC = 0.5
are compared with those returned by `wilcox.test()`.


```{r, simulation3}
# #############################################################################
# #
# # Simulation 3: confidence interval for pseudomedian derived under H0: AUC = 0.5   
# #               MC study of N = 500 replicas
# #               x ~ rnorm(300, 0,1)
# #               y ~ rlaplace(300, 0,1)
# #
# #############################################################################
#
#
#
# This simulation takes long time to complete
# N <- 500
# n_test <- 300
#
# set.seed(123L)
# wmw_ci = wt_ci = list(N)
# eauc = pseudomed = numeric(N)
# for (i in 1:N) {
#  #
#  x_test <- rnorm(n_test, 0, 1)
#  y_test <- VGAM::rlaplace(n_test, 0, 1)
#
#  wmw_test <- pseudomedian_ci(x_test, y_test, conf.level = 0.95)
#  wmw_ci[[i]] = wmw_test$conf.int
#  wt_test <- wilcox.test(x_test, y_test, conf.int = TRUE)
#  wt_ci[[i]] = wt_test$conf.int
#  eauc[i] = wt_test$statistic/(n_test*n_test)
#  pseudomed[i] = as.numeric(wt_test$estimate)
#  #  
# }
#
#
data(simulation3)  # List of wmw_ci, wt_ci, eauc, pseudomedian
# 
# Average across MC of confidence intervals obtained under H0: AUC=0.5
colMeans(simulation3$wmw_ci)
# Average across MC of confidence intervals from wilcox.test()
colMeans(simulation3$wt_ci)
# 
# Average across MC of eAUC
mean(simulation3$eauc)
 
# Coverage
length(which((simulation3$wmw_ci[,1] < 0) & (simulation3$wmw_ci[,2] > 0)))
length(which((simulation3$wt_ci[,1] < 0) & (simulation3$wt_ci[,2] > 0)))

# Mean pseudomedian
mean(simulation3$pseudomed)
```


## Example 1

Real data analyzed by WMW test of no group discrimination. 

```{r example1}
data(gemR::MS)
da <- MS

# preparing data frame
class(da$proteins) <- setdiff(class(da$proteins), "AsIs")
df <- as.data.frame(da$proteins)
df$MS <- da$MS
```

### Test of no group discrimination

```{r Ex1_discrimination}
wmd <- wmw_test(P19099 ~ MS, data = df, ref_level = 'no')
wmd

```

```{r plot_ex1, fig.width=5, fig.height=5, include=T, fig.align='center', fig.show='hold', out.width = "50%", echo = FALSE}
plot(wmd)
```


## Example 2

Synthetic data illustrating the special case of location shift assumption.

```{r example2}
data(Ex2)
da <- Ex2

# WMW test
wmd <- wmw_test(y ~ group, data = da, ref_level = 'control')
wmd
```

```{r ROC_example2, fig.width=5, fig.height=5, include=T, fig.align='center', fig.show='hold', out.width = "50%", echo = FALSE}
plot(wmd)
```

### Check location-shift assumption with EDA

```{r quadruplot_Ex2, fig.width=11, fig.height=8, include=T, fig.align='center', fig.show='hold', out.width = "100%", echo = FALSE}
# Check location-shift assumption with EDA
qp <- quadruplot(y ~ group, data = da, ref_level = 'control')
qp
# => location-shift assumption not tenable
```
location-shift assumption not tenable.


Erroneous use of location-shift special case of WMW would falsely conclude significant median difference despite identical medians.
```{r plot_example2, fig.width=10, fig.height=10, include=T, fig.align='center', fig.show='hold', out.width = "100%", echo = FALSE}
# Erroneous use of location-shift special case would falsely 
# conclude significant median difference despite identical medians
wml <- wmw_test(y ~ group, data = da, special_case = TRUE,
                 ref_level = 'control')
wml
```

Indeed, the medians are essentially the same:
```{r meds}
median(da$y[da$group == 'case'])
median(da$y[da$group == 'control'])
```


## Example 3

WMW applied to another real-life data set.

WARNING: the data contain ties. Current version of wmwAUC package does not take
ties into account.


```{r Ex3}
data(wesdr)
da = wesdr
da$ret = as.factor(da$ret)
# WMW 
wmd <- wmw_test(bmi ~ ret, data = da, ref_level = '0')
wmd
```

```{r roc_Ex3, fig.width=5, fig.height=5, include=T, fig.align='center', fig.show='hold', out.width = "50%", echo = FALSE}
plot(wmd)
```

### EDA to assess location shift assumption validity

```{r quadruplot_Ex3, fig.width=11, fig.height=8, include=T, fig.align='center', fig.show='hold', out.width = "100%", echo = FALSE}
# EDA to assess location shift assumption validity
qp <- quadruplot(bmi ~ ret, data = da, ref_level = '0')
qp
# => location shift assumption is tenable
```
hence, location shift assumption is tenable.

### Special case of WMW test
```{r Ex3_contd}
suppressWarnings({ # ties in data
wml <- wmw_test(bmi ~ ret, data = da, ref_level = '0', 
                 ci_method = 'boot', special_case = TRUE)
})                 
wml
```

Plot 
```{r plot_ex3, fig.width=11, fig.height=5, include=T, fig.align='center', fig.show='hold', out.width = "100%", echo = FALSE}
plot(wml)
```

# Acknowledgements

AI-assisted code generation via Claude Pro by Anthropic was used in development. 
All generated content was verified, tested, and enhanced by the package author.

# References


