---
output: github_document
title: 'wmwAUC Test of No Group Discrimination'
bibliography: inst/refs.bib
link-citations: true
editor_options: 
  markdown: 
    wrap: 72
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  warning = FALSE,
  message = FALSE
)
library('wmwAUC')
library('gemR')
library('gss')
#
# 
```

# wmwAUC

<!-- badges: start -->

[![CRAN
status](https://www.r-pkg.org/badges/version/wmwAUC)](https://cran.r-project.org/package=wmwAUC)
[![CRAN
downloads](https://cranlogs.r-pkg.org/badges/grand-total/wmwAUC)](https://cran.r-project.org/package=wmwAUC)

<!-- badges: end -->

The classical Wilcoxon-Mann-Whitney (WMW) test is calibrated under the
null hypothesis $\mathrm{H_0\colon F = G}$. However, the WMW test
statistic, when standardized, is the empirical AUC (eAUC), and hence the
test in fact assesses the set $\{\mathrm{(F,G)\colon AUC = 0.5}\}$ --
which does not match $\mathrm{H_0\colon F = G}$. This incoherence causes
that WMW test provides erroneous inferences.

Through Monte Carlo analysis of zero-mean heteroskedastic Gaussians and
corresponding asymptotic theory, we demonstrate this mismatch directly:
the eAUC concentrates on 0.5 even when $\mathrm{F \neq G}$, showing that
the classical test's calibration cannot be tracking distributional
equality. Moreover, the traditional alternative hypothesis of stochastic
dominance is unnecessarily restrictive; the WMW statistic is consistent
against the broader alternative $\mathrm{H_1\colon AUC \neq 0.5}$, as
established by Van Dantzig (1951).

**wmwAUC** implements a properly calibrated test of
$\mathrm{H_0\colon AUC = 0.5}$ based on the WMW statistic; for details,
see [@grendar2025wmw]. The package provides both asymptotic inference
and two finite-sample bias-correction methods:

-   *Exact Unbiased* (EU) Method: Universal approach handling data with
    arbitrary tie patterns through the mid-rank kernel and exact
    finite-sample unbiased variance estimation from Hoeffding
    decomposition theory. Reduces correctly to the continuous case when
    no ties are present.

-   *Bias-Corrected* (BC) Method: Alternative using individual 
    placement-component bias correction with $O(n^{-1})$
    finite-sample corrections and Welch-Satterthwaite degrees of  
    freedom. Derived for continuous data (Sect. 4 of the paper); the
    implementation extends this to ties via the same mid-rank kernel as EU,
    with conservative behavior confirmed by simulation to persist under ties.

The EU method serves as the default implementation, providing:

-   Universal applicability (handles any data type - continuous,
    discrete, or mixed)

-   Exact finite-sample unbiasedness (not asymptotic approximation)

-   Theoretically principled tie handling through mid-rank kernel

-   Retained real power at small sample sizes where BC's power can be
    close to zero

BC remains available for users who specifically want its more
conservative behavior.

Key functions include:

-   `wmwAUC_test()`: Main testing function, defaulting to the EU method,
    with an option to select BC instead

-   `wmwAUC_pvalue_BC()`: WMW AUC p-values via the BC method

-   `wmwAUC_pvalue_EU()`: WMW AUC p-values via the EU method, valid for
    any tie pattern

-   `wmwAUC_pseudomedian_ci()`: Confidence interval for the
    AUC-equalizing shift (pseudomedian), obtained by inverting the test

(The old names `wmw_test()`, `wmw_pvalue()`, `wmw_pvalue_ties()`, and
`pseudomedian_ci()` are not retained as functional aliases. Calling any of
them raises an informative error naming the current function to use instead.)


## Installation

You can install the development version of wmwAUC using

``` r
devtools::install_github('grendar/wmwAUC')
```

## Simulation 1

Consider the setting of two zero-mean different-scale gaussians. Then
the traditional $\mathrm{H_0\colon F = G}$ of the classical WMW test is
false and $\mathrm{H_0\colon F \neq G}$ holds.

The Monte Carlo simulation demonstrates that the normalized test
statistic $U/(n_1n_2)$ which is just eAUC, concentrates asymptotically
on 0.5 - the value expected under a true null hypothesis.

If the classical WMW test tested distributional equality, the test
statistic should not concentrate on its null value when distributions
clearly differ.

Also note that under $\mathrm{H_0\colon F \neq G}$, p-values should
concentrate near zero, yet the observed distribution is nearly uniform
with a slightly elevated first bins, consistent with testing a true null
hypothesis ($\mathrm{H_0\colon AUC = 0.5}$) using miscalibrated variance
estimation.

```{r simulation1}
#############################################################################
#
# Simulation 1: H0: F=G is erroneously too broad
#
#############################################################################
#
# This simulation takes several minutes to complete
# N = 10000
# n = 1000
# set.seed(123L)
# pval_wt = pval_wmwAUC = eauc = numeric(N)
# for (i in 1:N) {
#
#   x = rnorm(n, sd = 0.1)
#   y = rnorm(n, sd = 3)
#   # wilcox.test() of H0: F = G
#   wt = wilcox.test(x, y)
#   pval_wt[i] = wt$p.value
#   # wmwAUC_test() of H0: AUC = 0.5
#   pval_wmwAUC[i] = wmwAUC_pvalue_BC(x, y)
#   # eAUC
#   eauc[i] = wt$statistic/(n*n)
#   #
# }
data(simulation1)  # List eauc, pval_wt, pval_wmwAUC
#
```

```{r, hist_sim1_1, fig.width=5, fig.height=5, include=T, fig.align='center', fig.show='hold', out.width = "50%", echo = FALSE}
hist(simulation1$eauc)
```

Empirical AUC centered at 0.5 despite $\mathrm{F \neq G}$.

```{r, hist_sim1_2, fig.width=5, fig.height=5, include=T, fig.align='center', fig.show='hold', out.width = "50%", echo = FALSE}
hist(simulation1$pval_wt) 
```

Traditional p-values under $\mathrm{H_1}$ should concentrate near 0.

```{r, hist_sim1_3, fig.width=5, fig.height=5, include=T, fig.align='center', fig.show='hold', out.width = "50%", echo = FALSE}
# Correct p-values (correctly testing AUC = 0.5)  
hist(simulation1$pval_wmwAUC)
```

Correct p-values for testing $\mathrm{H_0\colon AUC = 0.5}$.

## Simulation 2

The two zero-mean different-scale gaussians setting does not satisfy the
traditional $\mathrm{H_1}$ of the stochastic dominance. But, as proved
by Van Dantzig in 1951, the WMW statistic is consistent for the broader
$\mathrm{H_1\colon AUC \neq 0.5}$.

```{r simulation2}
#############################################################################
#
# Simulation 2: H1: F stoch. dominates G is too narrow
#               WMW is consistent for broader H1: AUC != 0.5   
#
#############################################################################
#
#
# This simulation takes several minutes to complete
# N = 10000
# n = 1000
# set.seed(123L)
# pval_wt = pval_wmwAUC = eauc = numeric(N)
# for (i in 1:N) {
# #
#   # gaussians with different location and scale
#   # does not satisfy stochastic dominance
#   x = rnorm(n, 0, sd = 0.1)
#   y = rnorm(n, 0.5, sd = 3)
#   # wilcox.test H0: F = G vs H1: (F stochastically dominates G) OR (G stochastically dominates F)
#   wt = wilcox.test(x, y)
#   pval_wt[i] = wt$p.value
#   # wmwAUC_test H0: AUC = 0.5 vs H1: AUC neq 0.5
#   pval_wmwAUC[i] = wmwAUC_pvalue_BC(x, y)
#   # eAUC
#   eauc[i] = wt$statistic/(n*n)
# #
# }
data(simulation2)  # List of eauc, pval_wt, pval_wmwAUC
# WMW detects broader alternatives than traditional stochastic dominance
```

```{r,hist_sim2, fig.width=5, fig.height=5, include=T, fig.align='center', fig.show='hold', out.width = "50%", echo = FALSE}
hist(simulation2$eauc)
hist(simulation2$pval_wmwAUC)
hist(simulation2$pval_wt)

```

## Example 1

Real data analyzed by wmwAUC test of no group discrimination.

```{r example1}
data(gemR::MS)
da <- MS

# preparing data frame
class(da$proteins) <- setdiff(class(da$proteins), "AsIs")
df <- as.data.frame(da$proteins)
df$MS <- da$MS
```

### Test of no group discrimination

```{r Ex1_discrimination}
wmd <- wmwAUC_test(P19099 ~ MS, data = df, ref_level = 'no')
wmd

```

```{r plot_ex1, fig.width=10, fig.height=5, include=T, fig.align='center', fig.show='hold', out.width = "75%", echo = FALSE}
plot(wmd)
```

### AUC-equalizing shift (pseudomedian)

```{r Ex1_pseudomedian}
set.seed(123L)
wmd_pm <- wmwAUC_test(P19099 ~ MS, data = df, ref_level = 'no', pseudomedian = TRUE)
wmd_pm
```

# Acknowledgements

AI-assisted code generation via Claude Pro by Anthropic was used in
development. All generated content was verified, tested, and enhanced by
the package author.

# References
